Hockey's Most Controversial Statistic: An Analysis of the Effectiveness of the Plus-Minus Statistic

### 2024-12-06 response/response.tex

\documentclass[12pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage{natbib}
\usepackage{listings}
\usepackage{rotating, graphicx}
\usepackage{booktabs, natbib}
\usepackage[hyphens]{url}
% \usepackage [english]{babel}
\usepackage{amsmath, amsbsy, amsthm, epsfig, epsf, psfrag, graphicx,
  amssymb, enumerate, bm}
\usepackage{enumitem}

\usepackage{color}

\newcommand{\jy}[1]{\textcolor{red}{JY #1}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}

\usepackage[colorlinks=true, urlcolor=blue, linkcolor=blue, citecolor=blue]{hyperref}

\newenvironment{comment}%
{\begin{quotation}\noindent\small\it\ignorespaces%
  }{\end{quotation}}

\begin{document}

\begin{center}
  {\Large\bf Response Letter to Reviewers' Comments}
\end{center}

\section{Summary}

I would like to that Lucy Liu for her thoughtful, constructive
comments. The manuscript has been revised accordingly with the
following major changes:

\begin{enumerate}
\item
  Following the reviewer's comments on expanding the literature
  review, a new paragraph was added to the introduction that
  expanded on other works that have been done. This involved reviewing
  alternatives to the plus-minus and works that other people
  within the sports statistics world have done to address this.
  This also included addressing some downsides to these works and
  why this project is necessary.

\item
  In response to the reviewer's comment on clearly stating the 
  hypothesis, the hypothesis was added to the last paragraph of the 
  introduction that asked the over-arching question that is being
  addressed in the paper. This set up the work that is being done
  in the paper and what the readers can hope to take from this work.

\item
  The reasons that each of the models used were expanded. The need
  for the assumptions to varified was added to the methods section.
  The reviewer did make the suggestion of adding the plots to varify 
  that the assumptions were satisfied. While I did make plots to make
  sure all assumptions were varified, they were not included in the final
  paper due to space concerns. Also, I did not want to take attention
  away from the discussion at hand. There were additions made to the 
  ridge regression section explaining why ridge was chosen over other 
  regularization techniques.

\item 
  An extra paragraph was added to the end of the discussion to explain
  future directions. The reason for this study was re-iterated as well as
  explaining how it can be used in the NHL. Additional comments were added
  to explain what can be done in the future based on this study and the
  work of other people within the field.

\end{enumerate}

Point-by-point responses are as follows, with the reviewer/editor's comments in {\it italic}.

\section{Response to Each Comment}

\begin{comment}
  The literature review could be expanded to include additional studies that 
  have examined the plus-minus statistic and its alternatives. This could provide 
  a better foundation for the reasoning behind this study and highlight gaps
  that you hope to fill with your current research.
\end{comment}

This was a good note to improve the paper. While I initially mentioned Brian MacDonald's 
work and impliedthat other people had proposed alternatives to the plus-minus, I did not 
include enough detail. Including more literature helps the reader to have a better understanding
of what work is already being done. It also allows me to explain why my study is 
important and what it contributed to this conversation. A paragraph better explaining 
the literature review was added as a fourth paragraph in the introduction.

\begin{comment}
  The paper evaluates the effectiveness of the plus-minus statistic compared 
  to other metrics like Corsi and Fenwick. While the objectives are clear, the 
  hypotheses are not explicitly stated. Clearly state the hypotheses at the end 
  of the introduction.
\end{comment}

In the first draft, the overall goal of the project was stated as well as the 
questions that would be answered in the paper. At the request of the reviewer,
the hypothesis of the project was stated in the second to last sentence of the
introduction. Hopefully this helps the readers to understand the over-arching 
question that is being answered in this project.

\begin{comment}
  Although each method is described individually, the reason behind selecting these 
  specific methods (Ex.why ridge regression and not another regularization technique) 
  is not fully explained. Add a brief justification for each method and explain why 
  alternatives were not selected.

  Additionally, clearly outline the assumptions associated with each statistical 
  method and discuss how these assumptions are checked and addressed within the 
  analysis. Include diagnostic tests and results that assess whether the assumptions 
  of each statistical method are satisfied. For example, provide residual plots 
  or normality tests for mixed-effects models.
\end{comment}

\begin{comment}
   The discussion section is missing future directions. Discuss what comes next after 
   this study, or some possible new ideas or improvements to explore. Further elaborate 
   on the implications of these findings for the NHL.
\end{comment}

\bibliographystyle{chicago}
\bibliography{growth}

%\atColsBreak{\pagediscards}
\end{document}

### 2024-12-06 response/response.tex

In the first draft, the overall goal of the project was stated as well as the 
questions that would be answered in the paper. At the request of the reviewer,
the hypothesis of the project was stated in the second to last sentence of the
introduction. Hopefully this helps the readers to understand the over-arching 
question that is being answered in this project.

\begin{comment}
  Although each method is described individually, the reason behind selecting these 
  specific methods (Ex.why ridge regression and not another regularization technique) 
  is not fully explained. Add a brief justification for each method and explain why 
  alternatives were not selected.

  Additionally, clearly outline the assumptions associated with each statistical 
  method and discuss how these assumptions are checked and addressed within the 
  analysis. Include diagnostic tests and results that assess whether the assumptions 
  of each statistical method are satisfied. For example, provide residual plots 
  or normality tests for mixed-effects models.
\end{comment}

While in the first draft, the reason that each method was being used was explained,
the reviewer is correct that there is a need for additional explanation of why each
model works in the given context. A few sentences were added to the begining of the 
Ridge Regression section in Methods to explain how the models were used. Additionally,
the assumptions of the models needed to be addressed. While I did test that the 
assumptions were not violated, images were not included. Due to the length of the paper,
I did not want to include images to keep the paper from being incredibly long. Also, I felt
that it would not benefit the discussion to include a longer sectiona about verifying the
models' assumptions. A few sentences were added to the end of each of the models' sections 
to verify that the assumptions were met.

@@

\begin{comment}
   The discussion section is missing future directions. Discuss what comes next after 
   this study, or some possible new ideas or improvements to explore. Further elaborate 
   on the implications of these findings for the NHL.
\end{comment}

The reviewer is correct, the paper needs a better ending that encapsulates the work 
done. The initial last papragraph explained what was accomplished in this study and
breifly mentioned the implications of this work but failed to remind the readers why
it is important and what is can be done moving forward. An extra paragraph was added 
to the end of the Discussion section to explain why this study was important, what can 
be done with the information in this paper, and what next steps can be taken.

### 2024-12-06 manuscript/statspaper.tex
added based on comments

While the plus-minus statistic is a great idea in theory, it does not come without its weaknesses. The biggest drawback to this
statistic is its issues with independence. Many ice hockey players tend to play on "lines", meaning the same three 
forwards tend to play together and the same two defensemen usually play together. As a result, the performance of one player is highly dependent 
on the performance of their linemates. In "A Regression-based Adjusted Plus-Minus Statistic for NHL Players" \cite{Macdonald_2011}, Brian
MacDonald provides a perfect example of this with the Henrik and Daniel Sedin, Swedish twins who played together for the Vancouver Canucks. 
"Daniel spent 92\% of his playing time with Henrik, the highest percentage of any other player combination where both players have
played over 700 minutes. Because of this high collinearity between the twins, it is difficult to separate the individual effect that each 
player has on the net goals scored on the ice."\cite{Macdonald_2012} Many critics of the plus-minus argue that plus-minus is more of a team 
statistic, since it is heavily influenced by team dynamics rather than individual contributions. Furthermore, numerous confounding variables, 
such as the quality of the opponent and situational factors, make the calculation less reliable. As a result of these criticisms, alternatives 
to the plus-minus statistic have been developed, such as Corsi or Fenwick, which is calculated using shot attempts rather than goals. Brian 
MacDonald introduces an adjusted plus-minus statistic that is based on weighted least squares regression\cite{Macdonald_2012}.

The goal of this project will be to shed light on this controversial statistic and determine if plus-minus is truly an effective predictor
to individual contribution, specifically in the National Hockey League (henceforth refered to as the NHL). Is the plus-minus a good evaluator 
of offensive and defensive output? Is the plus-minus more reflective of how well a player is doing than how well a player is performing? 
Are there better alternatives to evaluating individual player contribution? These are the questions that will be answered from this analysis.

@@

Due to the presence of correlation among variables, ridge regression was performed in order to identify how plus/minus can be employed 
to assess offensive and defensive contribution. It was also used to determine if other advanced metrics Corsi or Fenwick have the same, 
better, or worse predictive abilities compared to plus-minus. Ridge regression, a form of regularized linear regression, is beneficial 
in handling datasets where predictor variables are highly interrelated, as is the case with advanced hockey metrics. Ridge regression 
estimates the contribution of metrics like Corsi, Fenwick, and scoring chances while controlling for their interdependencies. This 
applies to hockey metrics like Corsi, Fenwick, scoring chances, etc. since these statistics often interact or overlap in measuring aspects 
of performance. Ridge regression is appropriate because the data satisfies the model assumptions of independence, homoscedasticity, 
and linearity. It is ideal for this project because it effectively manages multicollinearity, retains all variables for interpretability, 
and improves predictive accuracy. Other regularization techniques like Lasso or Elastic Net are better suited for sparse models or when 
variable selection is a priority, which is not the project's focus.

@@

In order for fans, players, and sports management to have a better understanding of what make a player "effective", you have to 
summarize and quantify a large number of variables into one condensed number. That is, understandably, a tall order. While the plus-minus
lays the ground-work for getting a better understanding of player effectiveness, there is more at play. General managers of sports teams
often want to make the best team possible, getting players that will have a positive impact. Often times, people put a lot of emphasis on
direct offensive output, such as points, to determine a players contribution. But there is so much more to consider, such as defensive skills,
generating chances to score, etc, while still accounting for team effects. While plus-minus may not be the solution, the next step in the 
world of sports statistics is to find a way to quantify the concept of "player effectiveness".

>>>>

While the plus-minus statistic is a great idea in theory, it does not come without its weaknesses. The biggest drawback to this
statistic is its issues with independence. Many ice hockey players tend to play on "lines", meaning the same three 
forwards tend to play together and the same two defensemen usually play together. As a result, the performance of one player is highly dependent 
on the performance of their linemates. In "A Regression-based Adjusted Plus-Minus Statistic for NHL Players" \cite{Macdonald_2011}, Brian
MacDonald provides a perfect example of this with the Henrik and Daniel Sedin, Swedish twins who played together for the Vancouver Canucks. 
"Daniel spent 92\% of his playing time with Henrik, the highest percentage of any other player combination where both players have
played over 700 minutes. Because of this high collinearity between the twins, it is difficult to separate the individual effect that each 
player has on the net goals scored on the ice."\cite{Macdonald_2012} Many critics of the plus-minus argue that plus-minus is more of a team 
statistic, since it is heavily influenced by team dynamics rather than individual contributions. Furthermore, numerous confounding variables, 
such as the quality of the opponent and situational factors, make the calculation less reliable.

As a result of these criticisms, alternatives to the plus-minus statistic have been developed. Many of these alternatives are either different 
metrics or adjusted plus-minus\cite{plus_minus_rating}. Two popular examples are Corsi or Fenwick, which are calculated using shot attempts 
rather than goals. Another popular metric is expected goals, which is based on the probability of a goal being scored depending on the shot.
Brian MacDonald introduced an adjusted plus-minus statistic that is based on weighted least squares regression\cite{Macdonald_2012}. However,
many of these alternatives are more complicated to calculate, which can be confusing to the average fan. Due to the fact that plus-minus
is easy to interpret and calculate, it has remained in use despite its controversial background.

The goal of this project will be to shed light on this controversial statistic and determine if plus-minus is truly an effective predictor
to individual contribution, specifically in the National Hockey League (henceforth refered to as the NHL). Is the plus-minus a good evaluator 
of offensive and defensive output? Is the plus-minus more reflective of how well a player is doing than how well a player is performing? 
Are there better alternatives to evaluating individual player contribution? Is the plus-minus an effective predictor to individual contribution? 
These are the questions that will be answered from this analysis.

@@

Ridge regression is used in this project to evaluate the relationship between various performance metrics and plus-minus, while addressing 
multicollinearity among predictor variables. This regularization technique is particularly valuable in this context because many hockey 
performance metrics, such as Scoring Chances For percentage (SCF\%) and Goals For percentage (GF\%), are often highly correlated.
Due to the presence of correlation among variables, ridge regression was performed in order to identify how plus/minus can be employed 
to assess offensive and defensive contribution. It was also used to determine if other advanced metrics Corsi or Fenwick have the same, 
better, or worse predictive abilities compared to plus-minus. Ridge regression, a form of regularized linear regression, is beneficial 
in handling datasets where predictor variables are highly interrelated, as is the case with advanced hockey metrics. Ridge regression 
estimates the contribution of metrics like Corsi, Fenwick, and scoring chances while controlling for their interdependencies. This 
applies to hockey metrics like Corsi, Fenwick, scoring chances, etc. since these statistics often interact or overlap in measuring aspects 
of performance. Ridge regression is appropriate because the data satisfies the model assumptions of independence, homoscedasticity, 
and linearity. It is ideal for this project because it effectively manages multicollinearity, retains all variables for interpretability, 
and improves predictive accuracy. Other regularization techniques like Lasso or Elastic Net are better suited for sparse models or when 
variable selection is a priority, which is not the project's focus.

@@

In order for fans, players, and sports management to have a better understanding of what make a player "effective", you have to 
summarize and quantify a large number of variables into one condensed number. That is, understandably, a tall order. While the plus-minus
lays the ground-work for getting a better understanding of player effectiveness, there is more at play. General managers of sports teams
often want to make the best team possible, getting players that will have a positive impact. Often times, people put a lot of emphasis on
direct offensive output, such as points, to determine a players contribution. But there is so much more to consider, such as defensive skills,
generating chances to score, etc, while still accounting for team effects. Many have tried to do this, but lack the interpretability and 
simplicity of the plus-minus. While plus-minus may not be the solution, the next step in the world of sports statistics is to find a way to 
quantify the concept of "player effectiveness" in a way that includes the simplicity of the plus-minus without its controversial nature.
